---
title: "grooming_algorthim_test"
author: "david"
date: "2025-08-22"
output: html_document
---

##The goal of this code is to plot a transition matrix for fly grooming behavior, particularly grooming of the head, abdomen, thorax, and wings. We are going to just denote the number of transitions in each cell of the matrix in order to calculate the transition probability between behaviors (See Kalueff and Tuohimaa 2004 for an example).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r}
library(tidyverse)
library(psych)
library(pheatmap)
```

##creating and adding transition matrices 
```{r}
create_a_matrix <- function(x) {
  df <- read_csv(paste("C:/Users/user/Documents/Carleton College/Hoopfer lab/Summer 25/grooming algorthim/", x, sep = ""))
  
  #reading in a csv
  df <- df %>%
    select(Behavior, `Behavior type`, Time) #choosing only the relevant columns
  
  #separating start and stop 
  start_behavior <- df %>% 
    filter(row_number() %% 2 == 1)
  end_behavior <- df %>% 
    filter(row_number() %% 2 == 0)
  
  #rename the behavior names to 1-5
  start_behavior <- start_behavior %>%
    mutate(Behavior = case_when(Behavior == "Rubbing" ~ 1, 
                                Behavior == "Head" ~ 2, 
                                Behavior == "Abdomen" ~ 3,
                                Behavior == "Wing" ~ 4,
                                Behavior == "Thorax" ~ 5))
  end_behavior <- end_behavior %>%
    mutate(Behavior = case_when(Behavior == "Rubbing" ~ 1, 
                                Behavior == "Head" ~ 2, 
                                Behavior == "Abdomen" ~ 3,
                                Behavior == "Wing" ~ 4,
                                Behavior == "Thorax" ~ 5))
  
  #Creating empty matrix
  transition_matrix <- matrix(
    0, nrow = 5, ncol = 5,
    dimnames = list(c("rubbing","head","abdomen","wing","thorax"),
                    c("rubbing","head","abdomen","wing","thorax")))
  
  #Loop to add transitions into the matrix
  behavior_vector <- start_behavior$Behavior
  
  #for i in length of behavior_vector - 1, set row to value of i and set column to value of i + 1. i is the first start behavior, i+1 is the second behavior. Add 1 to the cell that corresponds to the coordinates of row and column in the matrix.
  for (i in 1:length(behavior_vector) - 1){
    row <- as.numeric(behavior_vector[i])
    column <- as.numeric(behavior_vector[i+1])
    transition_matrix[row, column] <- transition_matrix[row, column] + 1
  }
  #output 
  return(transition_matrix)
}

EH250416_SocIso_1_Scoring_CH1_DL <- create_a_matrix("EH250416_SocIso_1_Scoring_CH1_DL.csv")
EH250416_SocIso_1_Ch2_MRC <- create_a_matrix("EH250416_SocIso_1_Ch2_MRC.csv")
EH250416_SocIso_1_Scoring_CH3_DL <- create_a_matrix("EH250416_SocIso_1_Scoring_CH3_DL.csv")
EH250416_SocIso_1_Scoring_CH4_DL <- create_a_matrix("EH250416_SocIso_1_Scoring_CH4_DL.csv")
```

function to create transition probability
```{r}
calculate_transition_probabilities <- function(trans_matrix) {
  # row sums
  row_totals <- rowSums(trans_matrix)
  
  # take each value in the transition matrix by row and divide it by the corresponding row_totals value
  prob_matrix <- sweep(trans_matrix, 1, row_totals, FUN = "/")
  
  # replace NaN (from dividing by 0) with 0
  prob_matrix[is.na(prob_matrix)] <- 0
  
  # check each row sums to ~1 (within rounding)
  row_sums_check <- rowSums(prob_matrix)
  if (!all(abs(row_sums_check - 1) < 1e-8 | row_sums_check == 0)) {
    warning("Some rows do not sum to 1. Check your input matrix.")
  }
  
  return(prob_matrix)
}

tp_EH250416_SocIso_1_Scoring_CH1_DL <- calculate_transition_probabilities(EH250416_SocIso_1_Scoring_CH1_DL)
tp_EH250416_SocIso_1_Ch2_MRC <- calculate_transition_probabilities(EH250416_SocIso_1_Ch2_MRC)
tp_EH250416_SocIso_1_Scoring_CH3_DL <- calculate_transition_probabilities(EH250416_SocIso_1_Scoring_CH3_DL)
tp_EH250416_SocIso_1_Scoring_CH4_DL <- calculate_transition_probabilities(EH250416_SocIso_1_Scoring_CH4_DL)
```

function for averaging the probabilities 
```{r}
average_probabilities <- function(matrices_list) {
  # Check that all are matrices
  if (!all(sapply(matrices_list, is.matrix))) {
    stop("All elements in the list must be matrices")
  }
  
  # Sum them element-wise (by individual cells)
  summed <- Reduce("+", matrices_list)
  
  # Divide by the number of matrices to get the mean
  averaged <- summed / length(matrices_list)
  
  return(averaged)
}

matrices_list <- list(tp_EH250416_SocIso_1_Scoring_CH1_DL, 
                      tp_EH250416_SocIso_1_Ch2_MRC, 
                      tp_EH250416_SocIso_1_Scoring_CH3_DL,
                      tp_EH250416_SocIso_1_Scoring_CH4_DL)
test <- average_probabilities(matrices_list)
test
```

testing function to create heatmap
```{r}
# Create the heatmap (keeping original order)
pheatmap(
  test,
  color = colorRampPalette(c("blue", "white", "red"))(100),  # blue to red gradient
  cluster_rows = FALSE,  # keep original order of the data
  cluster_cols = FALSE,
  display_numbers = TRUE,  # show probabilities inside cells
  number_format = "%.2f",  # format values
  number_color = "white",
  main = "Transition Probability Heatmap",
  legend = TRUE
)
```

testing function to make violin plots
```{r}
#extract the probabilities from list of matrices
rubbing2rubbing <- sapply(matrices_list, function(x) x[1, 1])
rubbing2head <- sapply(matrices_list, function(x) x[1, 2])
rubbing2abdomen <- sapply(matrices_list, function(x) x[1, 3])
rubbing2wing <- sapply(matrices_list, function(x) x[1, 4])
rubbing2thorax <- sapply(matrices_list, function(x) x[1, 5])

#create data frame for plots
df_for_violin <- data.frame(rubbing2rubbing, rubbing2head, rubbing2abdomen, rubbing2wing, rubbing2thorax, condition = c(1:2)) 
df_for_violin <- df_for_violin %>%
  pivot_longer(
    cols = starts_with("rubbing"),  # which columns to pivot
    names_to = "Behavior Pair",       # new name for column with old column names
    values_to = "Probability"        # new name for values
  )

df_for_violin$condition <- as.factor(df_for_violin$condition)

#graphing violin plot 
test_plot <- ggplot(df_for_violin, aes(x = `Behavior Pair`, y = Probability, fill = condition)) +
  geom_violin(position = position_dodge(0.8)) +
  stat_summary(fun = mean, geom = "point", size = 2, color = "black",
               position = position_dodge(0.8), show.legend = FALSE) +
  theme_bw() +
  theme(axis.text = element_text(color = "black")) +
  labs(
    x = "Behavior Pair",
    y = "Probability"
  )
test_plot
```

